---
title: "Defossez et al OPLS"
output: html_notebook
---

```{r}
library(tidyverse)
library(here)
library(skimr)
library(janitor)
library(ropls)
library(chemhelper)
library(glue)
```


# Species level averages
Species-level averages are provided on Dryad.

```{r}
cardamine <-
  read_delim(here("data", "Defossez et al.", "all traits averages.csv"), delim = ";") %>%
  clean_names()
cardamine
```

- 'spp': the *Cardamine* species
- 'alt': ?
- 'light': ?
- 'moisture': ?
- 'ph': soil ph?
- 'nutrients': ?
- 'avg_alt': Altitude at center of distribution?
- 'fd_temperature': ?
- 'fd_,oisture': ?
- 'fd_radiation': ?
- 'wight': weight, total biomass (dry?)
- 'sla': specific leaf area. Ratio of 6mm diameter leaf disk area to its mass (mm^2^ mg^-1^)
- 'dmage': % damage where 100% means all leaves were completely eaten
- 'height': height
- 'photosy': chlorophyll content (SPAD)
- 'toughness': leaf toughness
- 'pieris':
- 'spod':
- 'c_n': carbon to nitrogen ratio
- 'gls_c': constitutive plant glucosinolate concentration
- 'gls_ja': Glucosinolate concentration in MeJA induced plants
- 'gls_ja_c': Difference between control and MeJA induced plants
- 'nb_peaks': ?
- 'voc_c': constitutive plant volatiles
- 'voc_ja': MeJA induced plant volatiles
- 'vocs_ja_c': difference between control and MeJA induced plant volatiles

```{r}
Xs <- c("wight", "sla", "toughness", "height", "photosy", "c_n", "gls_c", "gls_ja_c", "voc_c", "vocs_ja_c")
```

# PCA

```{r}
card.pca <- opls(cardamine %>% select(Xs), plotL = FALSE)
card.pca.data <- get_plotdata(card.pca)
ggplot(card.pca.data$scores, aes(x = p1, y = p2)) +
  geom_point(aes(color = cardamine$avg_alt)) +
  scale_color_viridis_c(option = "A") +
  labs(x = glue("P1 ({card.pca.data$var_explained[1,1] * 100}%)"),
       y = glue("P2 ({card.pca.data$var_explained[2,1] * 100}%)")) + theme_dark()
```


# PLSR
```{r}
opls(cardamine %>% select(Xs), cardamine$avg_alt, plotL = FALSE)
```
Good model!  Only one component, so I'll use OPLS

```{r}
card.opls <- opls(cardamine %>% select(Xs), cardamine$avg_alt, predI = 1, orthoI = NA)
plot_opls(card.opls) + scale_color_viridis_c(option = "A") + theme_dark()
```


# Compare variables
## PCA
```{r}
card.pca %>%
  get_loadings() %>% 
  arrange(desc(abs(p1)))
```
 so high altitude (negative on PC1) is associated with high constitutive glucosinolates, low SLA, high chlorophyll mostly
 
## OPLS
```{r}
card.opls %>% 
  get_VIP() %>% 
  arrange(desc(VIP))
```

ehh, lame.  pretty much the same conclusion.


# Individual data
I emailed Sergio Rasmann and got the un-summarized data, but it's confusing.

```{r}
raw <- read_delim(here("data", "Defossez et al.", "all_data_ind.csv"), ";") %>% clean_names()
raw
```

Within each population, they selected 8 plants, half of which were treated with JA. Some species have multiple populations.

- `species`: specific epithet, don't use--has wrong number of unique values
- `id`: unique plant ID, includes species abbrev., population number, induction treatment, and rep number
- `species_2`: Scientific name
- `treatment`: whether a plant was treated with jasmonic acid (JA) a control plan (C) or NA???
- `replicate`
- `treatment_nb`: don't use
- `leaveas_area_nub`
- `collection_date`
- `x_coord`
- `y_coord`
- `altitude`: Actual altitude where the plant was collected
- `avg_altitude`: Average altitude of the species from Swiss-wide occurance data
- `altitude_class`: Character representation of altitude, don't use.
- `av_weight`: weight of each plant
- `av_sla`: specific leaf area (ratio of 6mm diameter leaf disc area to its dry mass ($mm^2/mg$)
- `percent_damage`: ordinal percentage eaten (100% = all leaves gone)
- `av_height`: height of each plant
- `av_photosynt`: Actually chlorophyll content (SPAD)
- `av_toughness`: leaf toughness
- `weight_mg_cn`: sample weight for carbon/nitrogen analysis, ignore
- `nitrogen_percent`: %N content
- `carbon_percent`: %C content
- `cn`: Carbon to nitrogen ratio
- `hydrogen_percent`: ignore
- `gls_total`: glucosinolates in this specific plant
- `gls_c`: glucosinolates in this specific plant for control plants
- `gls_induced`: Mean `gls_total`for JA treated plants in this population - `gls_c`
- `voc_total`: VOCs in this specific plant
- `vocs_C`: VOCs in this specifc plant for control plants
- `vocs_induced`: Mean `voc_total` for JA treated plants in this population (species?) - `vocs_c`
- 

# Is population summary the same as species summary?

Maybe summarizing by population makes sense?

```{r}
raw %>%
  mutate(pop_id = str_extract(id, "\\d+")) %>% 
  select(id, pop_id, everything()) %>% 
  filter(!is.na(collection_date)) %>% 
  group_by(species_2, pop_id) %>% 
  count()
```
 Ummm....
 
 
Can I recapitulate species summary data??

```{r}
raw %>% 
  group_by(species_2) %>% 
  summarize(mean(cn, na.rm = TRUE))
```
```{r}
cardamine %>% select(spp, c_n)
```
NOPE!
