---
title: "Re-analysis of Muir et al. data"
author: "Eric R. Scott"
output: 
  html_notebook:
    code_folding: hide
    number_sections: yes
    theme: paper
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(ropls)
library(glue)
library(colorspace)
library(ggrepel)
source(here("R", "ropls_helpers.R"))
```

# Read in Data

I'm going to remove the outlier species, *S. juglandifolium* for the sake of this analysis

```{r}
tomato <- read_rds(here("data", "tomatoes.rds"))

#remove S.juglandifolium
tomato.filtered <-
  tomato %>% 
  filter(Spe != "jug") %>% 
  mutate(temp = temp * 0.1)
```


# PCA

```{r}
tom.pca2 <- opls(tomato.filtered %>% dplyr::select(lma.leaflet:Rdark), plotL = FALSE)
```

## Score plot:

```{r}
tom.pca2.data <- get_plotdata(tom.pca2)
ggplot(tom.pca2.data$scores, aes(x = p1, y = p2)) +
  geom_point(aes(fill = tomato.filtered$temp), shape = 21, size = 2) +
  scale_color_continuous_sequential("YlOrRd", aesthetics = "fill",
                                    name = "Temperature (ºC)") +
  labs(x = glue("PC1 ({tom.pca2.data$axis_stats[1,1] * 100}%)"),
       y = glue("PC2 ({tom.pca2.data$axis_stats[2,1] * 100}%)")) +
  theme_bw()
```

## Correlation plot

PC1 is the leaf economics spectrum

```{r}
pca.scores <- get_scores(tom.pca2)

pca.data <-
  tom.pca2@suppLs$xModelMN %>%
  as_tibble()

pca.cor.dat <-
  cor(pca.scores[2:3], pca.data, use = "pairwise.complete.obs") %>%
    t() %>%
    as_tibble(rownames = "variable") %>% 
  rowwise() %>% 
  mutate(distance = sqrt(sum((c(p1, p2) - c(0, 0))^2))) %>% 
  ungroup %>% 
  mutate(t = distance * sqrt((65-2)/(1-distance^2))) %>% 
  mutate(p.dist = pt(t, df = 65-2, lower.tail = FALSE)) %>% 
  mutate(p.adj = p.adjust(p.dist, method = "bonf"))
# pca.cor.dat
```
```{r}
pca.loading.plot <-
  ggplot(pca.cor.dat) +
  geom_segment(aes(x = 0, y = 0, xend = p1, yend = p2),
               arrow = arrow(length = unit(0.15, "cm"))) +
  # gghighlight(p.adj < 0.05, use_direct_label = FALSE) +
  geom_label_repel(aes(x = p1, y = p2, label = variable),
                   segment.alpha = 0.6, direction = "y", size = 2.7, 
                   min.segment.length = 0, force = 3) +
  theme_bw() +
  labs(x = "Correlation to PC1",
       y = "Correlation to PC2")
pca.loading.plot
```


# PCA regression

Temperature, but not precipitation, is significantly related to PC1 (i.e. the LES)

```{r}
pcr.df2 <-
  tom.pca2.data$scores %>% 
  add_column(temp = tomato.filtered$temp, precip = tomato.filtered$precip)
```

```{r}
pcr2.temp <- lm(temp ~ p1, data = pcr.df2)
pcr2.precip <- lm(precip ~ p1, data = pcr.df2)
pcr2.temp.a <- lm(temp ~ -1 + p1, data = pcr.df2)
pcr2.precip.a <- lm(precip ~ -1 + p1, data = pcr.df2)
summary(pcr2.temp.a)
summary(pcr2.precip.a)

car::Anova(pcr2.precip)

car::Anova(pcr2.temp)
```


# PLSR

Both precipitation **and** temperature are significantly related to leaf traits (but temperature has a much lower RMSEE).  They explain very little of the variation in the data (`R2X(cum)`), but a significant ammount of variation with the Y variables (`R2Y(cum)` and `pR2Y` and `pQ2`).  `Q2(cum)` is not much lower than R2Y indicating an absence of overfitting or a predictive power similar to the explanatory power.

```{r}
plsr2.precip <- 
  opls(tomato.filtered %>% dplyr::select(lma.leaflet:Rdark), tomato.filtered$precip, plotL = FALSE, permI = 1000)
# single component model

plsr2.temp <- 
  opls(tomato.filtered %>% dplyr::select(lma.leaflet:Rdark), tomato.filtered$temp, plotL = FALSE, permI = 1000)

```

View outputs in a table

```{r}
bind_rows("Precipitation ~ leaf traits" = plsr2.precip@summaryDF, "Temperature ~ leaf traits" = plsr2.temp@summaryDF, .id = "Model") %>% 
  select(-RMSEE, -pre, -ort)
```

# Compare Loadings

Table of PCA loadings and PLS axis loadings.

```{r}
precip.pls.loadings <- get_loadings(plsr2.precip) %>% rename(`Precipitation PLS` = p1)
precip.pls.vip <- get_VIP(plsr2.precip) %>% rename(`Precipitation VIP` = VIP)
temp.pls.loadings <- get_loadings(plsr2.temp) %>% rename(`Temperature PLS` = p1)
temp.pls.vip <- get_VIP(plsr2.temp) %>% rename(`Temperature VIP` = VIP)

pca.loadings <- get_loadings(tom.pca2) %>% rename(PC1 = p1, PC2 = p2)

loading.table <-
  full_join(pca.loadings, precip.pls.loadings) %>%
  full_join(precip.pls.vip) %>% 
  full_join(temp.pls.loadings) %>% 
  full_join(temp.pls.vip) %>% 
  mutate_if(is.numeric, ~round(., 3))

loading.table
write_csv(select(loading.table, PC1, `Precipitation PLS`, `Temperature PLS`),
          here("figs", "muir loading table.csv"))
```

It looks like maybe the Temperature PLS axis loadings are more similar to PC1, which makes sense because only the PCA regression for temperature was significant.


# PLSR with two response variables (both precip and temp)

Temp and precip aren't correlated, so this probably isn't useful.  That fact might be good justification for doing both separately though.

```{r}
ggplot(tomato.filtered, aes(x = precip, y = temp)) +
  geom_point()

cor.test(tomato.filtered$temp, tomato.filtered$precip)
```


Intermediate RMSEE. OK R2Y and Q2.  No p-value with this method.

```{r}
plsr.both <- 
  opls(tomato.filtered %>% dplyr::select(lma.leaflet:Rdark), tomato.filtered %>% dplyr::select(temp, precip) %>% as.matrix(), plotL = FALSE, permI = 200)
```

```{r}
getLoadingMN(plsr.both)

plsr.both@cMN
```

P1 is related to temperature, P2 is related to precipitation.  This supports doing separate PLS for precipitation and temperature
