---
title: "PCA vs. PLS-DA"
author: "Eric R. Scott"
output: html_notebook
---

This generates figure 2 showing the correlation structure and results from PCA-LR and PLS-DA for example datasets from the three scenarios.

**IMPORTANT:** These datasets were generated under R version 3.5.3.  The random number generator in R was updated in version 3.6.0.  For this to be fully reproducible, run with version < 3.6.0 or use `RNGversion("3.5.3")`.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ropls) #for opls(), which does pca and pls-da
library(cowplot) #for plot_grid()
theme_set(theme_gray())
library(here)
library(colorspace)
source(here("R", "ropls_plots.R"))
source(here("R", "ropls_helpers.R"))
source(here("R", "pcr.R"))
library(glue)
library(latex2exp)
```

```{r}
# Function to get brackets to play nicely with ggplot2. Taken from https://stackoverflow.com/questions/35633239/add-curly-braces-to-ggplot2-and-then-use-ggsave

library(pBrackets)
bracketsGrob <- function(...){
l <- list(...)
e <- new.env()
e$l <- l
  grid:::recordGrob(  {
    do.call(grid.brackets, l)
  }, e)
}
```

# Read in datasets
I'll read in the datasets from "3-generating-datasets.Rmd" and pick individual datasets to make figures

```{r}
null.list <- read_rds(here::here("data", "null.rds"))
control.list <- read_rds(here::here("data", "control.rds"))
needle.list <- read_rds(here::here("data", "needle.rds"))
```

Sample one dataset as an example

```{r}
set.seed(888)
null <- sample(null.list, 1)[[1]]
control <- sample(control.list, 1)[[1]]
needle <- sample(needle.list, 1)[[1]]
```

# Null

## Correlation heatmap for Null

```{r}
null.cor <-
  null %>% 
  select(-group) %>% 
  cor() %>%
  as_tibble(rownames = "row") %>% 
  gather(-row, key = col, value = cor)

null.cor.p <-
  ggplot(null.cor, aes(x = col, y = row, fill = cor)) +
  geom_tile() +
  scale_fill_continuous_diverging(palette = "Blue-Red 3", limits = c(-1,1)) +
  labs(x = NULL, y = NULL, subtitle = '"Null"', fill = "r") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "right")

# Replace axis labels with curly braces to improve readability:

b1 <- bracketsGrob(10/25, 0, 0.005, 0, h = 0.05)
b2 <- bracketsGrob(20/25, 0, 10/25, 0, h = 0.05)
b3 <- bracketsGrob(0.995, 0, 20/25, 0, h = 0.05)

bracket_labels <- tibble(x = c(5.5, 15.5, 23), #where to put labels on x-axis
                         y = rep(-1.75, 3), #how far below bracket to place labels
                         label = c("C1", "C2", "N")) #labels for groups of variables

null.cor.p <-
  null.cor.p +
  coord_cartesian(ylim = c(1,25), clip = "off") + #allows brackets to show up off plot
  theme(
    axis.text = element_blank(), #turns off axis labels
    axis.ticks = element_blank(), #turns off axis ticks
    plot.margin = unit(c(1,1,1.5,1), "lines") #padds margin to make room
  ) +
  annotation_custom(b1) +
  annotation_custom(b2) +
  annotation_custom(b3) +
  geom_text(data = bracket_labels, inherit.aes = FALSE, aes(x=x, y=y, label = label))

# null.cor.p
```


## PCA and PLS-DA for Null

```{r}
null.pcalr <- pca_lr(null, -group, group, CV = 7)
null.pca <- null.pcalr$pca
null.pls <- opls(select(null, -group), null$group, plotL = FALSE, permI = 500, predI = 2)
```

## Create plots for Null

```{r message=FALSE, warning=FALSE}
null.pca.p <-
  plot_pca(null.pca, null$group, annotate = "subtitle") +
  labs(title = NULL,
       subtitle =TeX(glue("$R^2_Y = {round(null.pcalr$mod.stats$R2Y, 3)}$; $p = {round(null.pcalr$mod.stats$p.value, 3)}$"))) +
  scale_color_discrete_qualitative(palette = "Dark 3")

null.pls.p <-
  plot_plsda(null.pls, annotate = "subtitle") +
  scale_color_discrete_qualitative(palette = "Dark 3")
```


# Needle in a haystack

## Correlation heatmap for needle-in-a-haystack

```{r}
needle.cor <-
  needle %>% 
  select(-group) %>% 
  cor() %>%
  as_tibble(rownames = "row") %>%  
  gather(-row, key = col, value = cor)

needle.cor.p <-
  ggplot(needle.cor, aes(x = col, y = row, fill = cor)) +
  geom_tile() +
  scale_fill_continuous_diverging(palette = "Blue-Red 3", limits = c(-1,1)) +
  labs(x = NULL, y = NULL, subtitle = '"Needle in a haystack"', fill = "r") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "right")

b1 <- bracketsGrob(10/25, 0, 0.005, 0, h = 0.05)
b2 <- bracketsGrob(20/25, 0, 10/25, 0, h = 0.05)
b3 <- bracketsGrob(0.995, 0, 20/25, 0, h = 0.05)

bracket_labels <- tibble(x = c(5.5, 15.5, 23), #where to put labels on x-axis
                         y = rep(-1.75, 3), #how far below bracket to place labels
                         label = c("C1", "C2", "D")) #labels for groups of variables

needle.cor.p <-
  needle.cor.p +
  coord_cartesian(ylim = c(1,25), clip = "off") + #allows brackets to show up off plot
  theme(
    axis.text = element_blank(), #turns off axis labels
    axis.ticks = element_blank(), #turns off axis ticks
    plot.margin = unit(c(1,1,1.5,1), "lines") #pads margin to make room
  ) +
  annotation_custom(b1) +
  annotation_custom(b2) +
  annotation_custom(b3) +
  geom_text(data = bracket_labels, inherit.aes = FALSE, aes(x=x, y=y, label = label))
```

## PCA and PLS-DA for needle-in-a-haystack

The PCA score plot still shows no real separation because the discriminating variables do not contribute much to the total covariation in the data.  The PLS-DA plot shows very strong separation, despite the differences between the groups being only due to 5 out of 25 variables.  The PLS-DA model performs well, with high R^2^ and Q^2^ values

```{r include=FALSE}
needle.pcalr <- pca_lr(needle, -group, group, CV = 7)
needle.pca <- needle.pcalr$pca
needle.pls <- opls(select(needle, -group), needle$group, plot = FALSE, predI = 2, permI = 500)
```

## Plots for needle-in-a-haystack

```{r message=FALSE, warning=FALSE}
needle.pca.p <-
  plot_pca(needle.pca, needle$group, annotate = "subtitle") +
  labs(title = NULL,
       subtitle =TeX(glue("$R^2_Y = {round(needle.pcalr$mod.stats$R2Y, 3)}$; $p = {round(needle.pcalr$mod.stats$p.value, 3)}$"))) +
  scale_color_discrete_qualitative(palette = "Dark 3")

needle.pls.p <-
  plot_plsda(needle.pls, annotate = "subtitle") +
  scale_color_discrete_qualitative(palette = "Dark 3")
```


# Control

## Correlation heatmap for Control

```{r}
control.cor <-
  control %>% 
  select(-group) %>% 
  cor() %>%
  as_tibble(rownames = "row") %>% 
  gather(-row, key = col, value = cor)

control.cor.p <-
  ggplot(control.cor, aes(x = col, y = row, fill = cor)) +
  geom_tile() +
  scale_fill_continuous_diverging(palette = "Blue-Red 3", limits = c(-1,1)) +
  labs(x = NULL, y = NULL, subtitle = '"Control"', fill = "r") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "top")

b1 <- bracketsGrob(5/25, 0, 0.005, 0, h = 0.05)
b2 <- bracketsGrob(10/25, 0, 5/25, 0, h = 0.05)
b3 <- bracketsGrob(15/25, 0, 10/25, 0, h = 0.05)
b4 <- bracketsGrob(0.995, 0, 15/25, 0, h = 0.05)

bracket_labels <- tibble(x = c(3, 8, 13, 20.5), #where to put labels on x-axis
                         y = rep(-1.75, 4), #how far below bracket to place labels
                         label = c("C", "D1", "D2", "N")) #labels for groups of variables
control.cor.p <-
  control.cor.p +
  coord_cartesian(ylim = c(1,25), clip = "off") + #allows brackets to show up off plot
  theme(
    axis.text = element_blank(), #turns off axis labels
    axis.ticks = element_blank(), #turns off axis ticks
    plot.margin = unit(c(1,1,1.5,1), "lines") #padds margin to make room
  ) +
  annotation_custom(b1) +
  annotation_custom(b2) +
  annotation_custom(b3) +
  annotation_custom(b4) +
  geom_text(data = bracket_labels, inherit.aes = FALSE, aes(x=x, y=y, label = label))
```

## PCA and PLS-DA for Control

```{r include=FALSE}
control.pcalr <- pca_lr(control, -group, group, CV = 7)
control.pca <- control.pcalr$pca
control.pls <- opls(select(control, -group), control$group, plot = FALSE, predI = 2, permI = 500)
```

## Plots for Control

```{r message=FALSE, warning=FALSE}
control.pca.p <-
  plot_pca(control.pca, control$group, annotate = "subtitle") +
  labs(title = NULL,
       subtitle =TeX(glue("$R^2_Y = 0.999$; $p < 0.001$"))) +
  scale_color_discrete_qualitative(palette = "Dark 3")

control.pls.p <-
  plot_plsda(control.pls, annotate = "subtitle") +
  scale_color_discrete_qualitative(palette = "Dark 3")
```


# Make figure

Put all the plots together to make a figure.

```{r}
fix_theme_cor <-
  theme(legend.position = "none",
        plot.subtitle = element_text(hjust = 0.5),
        plot.margin = unit(c(0.5,0.5,1.5,0.5), "lines"))

fix_theme_score <-
  theme(legend.position = "none",
        plot.margin = unit(c(0.5,0.5,0.5,0.5), "lines"))

color.bar <- get_legend(needle.cor.p)

plot_grid(null.cor.p + fix_theme_cor,
          control.cor.p + fix_theme_cor,
          needle.cor.p + fix_theme_cor,
          null.pca.p + fix_theme_score,
          control.pca.p + fix_theme_score,
          needle.pca.p + fix_theme_score,
          null.pls.p + fix_theme_score + labs(title = NULL),
          control.pls.p + fix_theme_score + labs(title = NULL),
          needle.pls.p + fix_theme_score + labs(title = NULL),
          nrow = 3, ncol = 3, align = "v", axis = "l", labels = "AUTO", label_x = 0.05) %>% 
  plot_grid(color.bar, align = "h", axis = "t", rel_widths = c(12,1)) %>%
  save_plot(here("figs", "three-scenarios.png"), ., ncol = 3, nrow = 3, base_height = 3)
```

