---
title: "PCA vs. PLS-DA"
author: "Eric R. Scott"
output: html_notebook
---
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ropls) #for opls(), which does pca and pls-da
library(cowplot) #for plot_grid()
theme_set(theme_gray())
library(here)
library(colorspace)
source(here("R", "ropls_plots.R"))
source(here("R", "ropls_helpers.R"))
```

# To-Do:

- Place a single copy of the correlation matrix legend in the top right or lower left of figure.
- Get figures to align without making margins huge and wasting a lot of space

# Read in datasets
I'll read in the datasets from "Generating Datasets.Rmd" and pick individual datasets to make figures

```{r}
null.list <- read_rds(here::here("data", "null.rds"))
control.list <- read_rds(here::here("data", "control.rds"))
needle.list <- read_rds(here::here("data", "needle.rds"))
```

Sample one dataset as an example
Note: the random number generator changed in R version 3.6 and will not result in the same plot

```{r}
# set.seed(999)
set.seed(888)
null <- sample(null.list, 1)[[1]]
control <- sample(control.list, 1)[[1]]
needle <- sample(needle.list, 1)[[1]]
```

# Null

## Correlation heatmap for Null

```{r}
null.cor <-
  null %>% 
  select(-group) %>% 
  cor() %>%
  as_tibble(rownames = "row") %>% 
  gather(-row, key = col, value = cor)

null.cor.p <-
  ggplot(null.cor, aes(x = col, y = row, fill = cor)) +
  geom_tile() +
  scale_fill_continuous_diverging(palette = "Blue-Red 3", limits = c(-1,1)) +
  labs(x = NULL, y = NULL, subtitle = '"Null"', fill = "r") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "right")
```


## PCA and PLS-DA for Null

```{r}
null.pca <- opls(select(null, -group), plotL = FALSE)
# null.pls <- opls(select(null, -group), null$group, plot = FALSE, permI = 500)
null.pls <- opls(select(null, -group), null$group, plotL = FALSE, permI = 500, predI = 2)
```

## Create plots for Null

```{r message=FALSE, warning=FALSE}
null.pca.p <-
  plot_pca(null.pca, null$group, annotate = "subtitle") +
  scale_color_discrete_qualitative(palette = "Dark 3")

null.pls.p <-
  plot_plsda(null.pls, annotate = "subtitle") +
  scale_color_discrete_qualitative(palette = "Dark 3")

plot_grid(null.cor.p,
          null.pca.p + theme(legend.position = "NULL"),
          null.pls.p + theme(legend.position = "NULL"), nrow = 1, ncol = 3) 
```
```{r}
# for lab meeting talk.  Delete before archiving
plot_grid(null.cor.p + theme(legend.position = "right"),
          NULL,
          null.pca.p + theme(legend.position = "NULL"),
          null.pls.p + theme(legend.position = "NULL"),
          nrow = 2, ncol = 2,
          rel_heights = c(0.7, 1)) %>% 
save_plot(filename = here("figs", "null plot.png"), plot = .,
          ncol = 2, nrow = 2, base_height = 3.5)
```

# Needle in a haystack

## Correlation heatmap for needle-in-a-haystack

```{r}
needle.cor <-
  needle %>% 
  select(-group) %>% 
  cor() %>%
  as_tibble(rownames = "row") %>%  
  gather(-row, key = col, value = cor)

needle.cor.p <-
  ggplot(needle.cor, aes(x = col, y = row, fill = cor)) +
  geom_tile() +
  scale_fill_continuous_diverging(palette = "Blue-Red 3", limits = c(-1,1)) +
  labs(x = NULL, y = NULL, subtitle = '"Needle in a haystack"', fill = "r") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "right")
```

## PCA and PLS-DA for needle-in-a-haystack

The PCA score plot still shows no real separation because the discriminating variables do not contribute much to the total covariation in the data.  The PLS-DA plot shows very strong separation, despite the differences between the groups being only due to 5 out of 25 variables.  The PLS-DA model performs well, with high R^2^ and Q^2^ values

```{r include=FALSE}
needle.pca <- opls(select(needle, -group), plot = FALSE)
needle.pls <- opls(select(needle, -group), needle$group, plot = FALSE, predI = 2, permI = 500)
```

## Plots for needle-in-a-haystack

```{r message=FALSE, warning=FALSE}
needle.pca.p <-
  plot_pca(needle.pca, needle$group, annotate = "subtitle") +
  scale_color_discrete_qualitative(palette = "Dark 3")

needle.pls.p <-
  plot_plsda(needle.pls, annotate = "subtitle") +
  scale_color_discrete_qualitative(palette = "Dark 3")

plot_grid(needle.cor.p,
          needle.pca.p + theme(legend.position = "NULL"),
          needle.pls.p + theme(legend.position = "NULL"),
          nrow = 1, ncol = 3)
```
```{r}
# for lab meeting talk.  Delete before archiving

plot_grid(needle.cor.p + theme(legend.position = "right"),
          NULL,
          needle.pca.p + theme(legend.position = "NULL"),
          needle.pls.p + theme(legend.position = "NULL"),
          nrow = 2, ncol = 2,
          rel_heights = c(0.7, 1)) %>% 
save_plot(filename = here("figs", "needle plot.png"), plot = .,
          ncol = 2, nrow = 2, base_height = 3.5)
```

# Control

## Correlation heatmap for Control

```{r}
control.cor <-
  control %>% 
  select(-group) %>% 
  cor() %>%
  as_tibble(rownames = "row") %>% 
  gather(-row, key = col, value = cor)

control.cor.p <-
  ggplot(control.cor, aes(x = col, y = row, fill = cor)) +
  geom_tile() +
  scale_fill_continuous_diverging(palette = "Blue-Red 3", limits = c(-1,1)) +
  labs(x = NULL, y = NULL, subtitle = '"Control"', fill = "r") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "top")
```

## PCA and PLS-DA for Control

```{r include=FALSE}
control.pca <- opls(select(control, -group), plot = FALSE)
control.pls <- opls(select(control, -group), control$group, plot = FALSE, predI = 2, permI = 500)
```

## Plots for Control

```{r message=FALSE, warning=FALSE}
control.pca.p <-
  plot_pca(control.pca, control$group, annotate = "subtitle") +
  scale_color_discrete_qualitative(palette = "Dark 3")

control.pls.p <-
  plot_plsda(control.pls, annotate = "subtitle") +
  scale_color_discrete_qualitative(palette = "Dark 3")

plot_grid(control.cor.p,
          control.pca.p + theme(legend.position = "NULL"),
          control.pls.p + theme(legend.position = "NULL"),
          nrow = 1, ncol = 3)
```
```{r}
# for lab meeting talk.  Delete before archiving

plot_grid(control.cor.p + theme(legend.position = "right"),
          NULL,
          control.pca.p + theme(legend.position = "NULL"),
          control.pls.p + theme(legend.position = "NULL"),
          nrow = 2, ncol = 2,
          rel_heights = c(0.7, 1)) %>% 
save_plot(filename = here("figs", "control plot.png"), plot = .,
          ncol = 2, nrow = 2, base_height = 3.5)
```

# Make figure

Put all the plots together to make a figure.

```{r}
fig1_new <-
  plot_grid(null.cor.p + theme(legend.position = "none"),
            null.pca.p + theme(legend.position = "none") + labs(title = NULL),
            null.pls.p + theme(legend.position = "none") + labs(title = NULL),
            control.cor.p + theme(legend.position = "none"),
            control.pca.p + theme(legend.position = "none") + labs(title = NULL),
            control.pls.p + theme(legend.position = "none") + labs(title = NULL),
            needle.cor.p + theme(legend.position = "none"),
            needle.pca.p + theme(legend.position = "none") + labs(title = NULL),
            needle.pls.p + theme(legend.position = "none") + labs(title = NULL),
            ncol = 3, nrow = 3,
            labels = "AUTO",
            align = "hv", axis = "bl")
# fig1_new

save_plot(here("figs", "fig1_new.png"), fig1_new, ncol = 3, nrow = 3)
```


### another possibility

```{r}
top.row <- plot_grid(null.cor.p + theme(legend.position = "none", plot.subtitle = element_text(hjust = 0.5)),
                     control.cor.p + theme(legend.position = "none", plot.subtitle = element_text(hjust = 0.5)),
                     needle.cor.p + theme(legend.position = "none", plot.subtitle = element_text(hjust = 0.5)),
                     nrow = 1,
                     labels = c("A", "B", "C"))
middle.row <- plot_grid(null.pca.p + theme(legend.position = "none") + labs(title = NULL),
                        control.pca.p + theme(legend.position = "none") + labs(title = NULL),
                        needle.pca.p + theme(legend.position = "none") + labs(title = NULL),
                        nrow = 1,
                        labels = c("D", "E", "F"))
bottom.row <- plot_grid(null.pls.p + theme(legend.position = "none") + labs(title = NULL),
                        control.pls.p + theme(legend.position = "none") + labs(title = NULL),
                        needle.pls.p + theme(legend.position = "none") + labs(title = NULL),
                        nrow = 1,
                        labels = c("G", "H", "I"))
color.bar <- get_legend(needle.cor.p)

top <- align_plots(top.row, color.bar, align = "h", axis = "t")

fig_alt <- plot_grid(top[[1]], middle.row, bottom.row, ncol = 1) %>% plot_grid(top[[2]], rel_widths = c(12, 1))
save_plot(here("figs", "fig1_alt.png"), fig_alt, ncol = 3, nrow = 3)
```

