---
title: "Simulating Datasets"
output: html_notebook
---
```{r message=FALSE, warning=FALSE}
library(iheatmapr)
# library(rlist)
library(chemhelper)
library(tidyverse)
# myseed <- runif(1, 1, 1000)
myseed <- 258
```

To Do:

- probs make more datasets than just 100.


It makes sense to make randomly generated datasets under my three scenarios **once** and then save them as .RDS files.  Then, I can do comparisons of PCA and PLS-DA, pick individual datasets to make figures out of, etc. using these saved files.

**For all datasets:**

 - p (number of variables) = 25
 - N (number of observations) = 20
 - var (variation parameter, passed to mvrnorm) = 1
 - Number of groups = 2
 - Number of datasets generated = 100
 
**Scenarios:**

1. **NULL**
    - no discriminating variables
    - 5 variables with cov = 0
    - two groups of 10 variables with cov = 0.5
  
2. **Needle in a haystack**
    - 5 variables that discriminate between groups with a difference in means of 2, cov = 0
    - two groups of 10 variables with cov = 0.5
  
3. **Red herring**
    - one group of 10 variables with d = 1 and cov = 0.5 (high variation, low discrimination)
    - one group of 10 variables with d = 2 and cov = 0 (low variation, high discrimination)
    - 5 variables with cov = 0
    
4. **Control**
    - 5 variables that discriminate between groups with difference in means of 2, cov = 0.5
    - 10 variables with cov = 0
    - 10 variables with cov = 0.2 (weakly covarying, but not discriminating.  Something for axis2 to pick up on?)

# 1. NULL

```{r}
set.seed(myseed)
null <-
  map(1:100,
      ~ sim_df(N = 20, n_groups = 2) %>% 
        sim_covar(5, var = 1, cov = 0, name = "N") %>% 
        sim_covar(10, var = 1, cov = 0.5, name = "C1") %>% 
        sim_covar(10, var = 1, cov = 0.5, name = "C2")
  ) %>%
  set_names(paste("null_df_", 1:100))
```

Inspect one:

```{r}
set.seed(myseed)
null[[sample(1:100, 1)]] %>%
  dplyr::select(-group) %>% 
  cor() %>% 
  iheatmap(row_labels = TRUE, col_labels = TRUE, name = "cor")
```

# 2. Needle in a haystack

```{r}
set.seed(myseed)
needle_in_haystack <-
  map(1:100,
      ~ sim_df(N = 20, n_groups = 2) %>% 
        sim_discr(5, var = 1, cov = 0, group_means = c(-1, 1), name = "D") %>% 
        sim_covar(10, var = 1, cov = 0.5, name = "C1") %>% 
        sim_covar(10, var = 1, cov = 0.5, name = "C2")
  ) %>%
  set_names(paste("needle_in_haystack_", 1:100))
```

Inspect one:

```{r}
needle_in_haystack[[sample(1:100, 1)]] %>%
  dplyr::select(-group) %>% 
  cor() %>% 
  iheatmap(row_labels = TRUE, col_labels = TRUE, name = "cor")
```

# 3. Red Herring

```{r}
set.seed(myseed)
red_herring <-
  map(1:100,
      ~ sim_df(N = 20, n_groups = 2) %>% 
        sim_covar(5, var = 1, cov = 0, name = "N") %>% 
        sim_discr(10, var = 1, cov = 0.5, group_means = c(-0.5, 0.5), name = "highcov") %>% 
        sim_discr(10, var = 1, cov = 0, group_means = c(-1, 1), name = "lowcov")
  ) %>%
  set_names(paste("red_herring_", 1:100))
```

Inspect one:

```{r}
red_herring[[sample(1:100, 1)]] %>%
  dplyr::select(-group) %>% 
  cor() %>% 
  iheatmap(row_labels = TRUE, col_labels = TRUE, name = "cor")
```

# 4. Control
```{r}
set.seed(myseed)
control <-
  map(1:100,
      ~ sim_df(N = 20, n_groups = 2) %>% 
        sim_discr(5, var = 1, cov = 0.5, group_means = c(-1, 1), name = "D1") %>% 
        sim_discr(5, var = 1, cov = 0.5, group_means = c(-1, 1), name = "D2") %>% 
        sim_covar(10, var = 1, cov = 0, name = "N") %>% 
        sim_covar(5, var = 1, cov = 0.5, name = "C")
  ) %>%
  set_names(paste("control_", 1:100))
```

inspect one:

```{r}
control[[sample(1:100, 1)]] %>%
  dplyr::select(-group) %>% 
  cor() %>% 
  iheatmap(row_labels = TRUE, col_labels = TRUE, name = "cor")
```


# Save 'em
```{r}
library(here)
write_rds(null, here("data", "null.rds"))
write_rds(needle_in_haystack, here("data", "needle.rds"))
write_rds(red_herring, here("data", "red herring.rds"))
write_rds(control, here("data", "control.rds"))
```

