---
title: "Mini-review"
author: "Eric R. Scott"
date: "2019-10-28"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
#for venn diagram
library(ggforce)
library(limma)
```

*Last compiled: `r Sys.Date()`*

# Mini-review methods

Journals searched:

- [Ecological Monographs](https://esajournals.onlinelibrary.wiley.com/journal/15577015)
- [Ecology](https://esajournals.onlinelibrary.wiley.com/journal/19399170)
- [Ecology Letters](https://onlinelibrary.wiley.com/journal/14610248)
- [Journal of Ecology](https://besjournals.onlinelibrary.wiley.com/journal/13652745)
- [Oecologia](https://www.springer.com/journal/442/)
- [Oikos](https://onlinelibrary.wiley.com/toc/16000706/current)


Search terms:

- first round:  PCA OR PLS OR "principal component" OR "partial least squares" OR "partial least square"
- second round: RDA or "redundancy analysis"


# Explore and create Table

To skip the wrangling above and just re-run these analyses:

```{r}
df_supp <- read_csv(here::here("data", "mini-review.csv"))
```

## How many papers total and broken down by journal?

```{r}
df_supp
nrow(df_supp)
count(df_supp, journal)
```

## how many papers use each method

```{r}
df_supp %>% 
  summarize(PCA = sum(!is.na(pca_use)), 
            PLS = sum(!is.na(pls_use)),
            RDA = sum(!is.na(rda_use)))
```


All unsupervised vs. all supervised (in numbers of publications)

```{r}
df_supp %>% 
  summarize(unsupervised = sum(!is.na(pca_use) | !is.na(other_unsupervised)),
            supervised = sum(!is.na(pls_use) | !is.na(rda_use) | !is.na(other_supervised)))
```

All unsupervised vs. all supervised (in numbers of applications.  That is, counting each method)

```{r}
df_supp %>% 
  summarize(unsupervised = sum(!is.na(pca_use), !is.na(other_unsupervised)),
            supervised = sum(!is.na(pls_use), !is.na(rda_use), !is.na(other_supervised)))
```


```{r}
df_venn <- 
  df_supp %>%
  select(pca_use, rda_use, pls_use) %>% 
  mutate_all(~ifelse(is.na(.), 0, 1))

vennCounts(df_venn) %>% vennDiagram()
```

Prettier version:


```{r}
vdc <- vennCounts(df_venn)
vdc
df.venn <- data.frame(x = c(0, 0.866, -0.866),
                      y = c(1, -0.5, -0.5),
                      labels = c('PLS', 'RDA', 'PCA'))
```


```{r}
class(vdc) <- 'matrix'
df.vdc <- as.data.frame(vdc)[-1,] %>%
  mutate(x = c(0, 1.2, 0.8, -1.2, -0.8, 0, 0),
         y = c(1.2, -0.6, 0.5, -0.6, 0.5, -1, 0))

ggplot(df.venn, aes(x0 = x, y0 = y, r = 1.5, fill = labels)) +
    geom_circle(alpha = .3, size = 1, colour = 'grey') +
      coord_fixed() +
  annotate("text", x = df.vdc$x, y = df.vdc$y, label = df.vdc$Counts, size = 5) +
        theme_void()
```


## How is PLS used?

```{r}
pls_use <-
  df_supp %>%
  filter(!is.na(pls_use)) %>% 
  mutate(pls_use = fct_infreq(pls_use)) %>% 
  count(pls_use) %>%
  rename("Use" = pls_use,
         "Number of publications" = n)
pls_use
```

## How is RDA used?

```{r}
rda_use <-
  df_supp %>%
  filter(!is.na(rda_use)) %>% 
  mutate(rda_use = fct_infreq(rda_use)) %>% 
  count(rda_use) %>%
  rename("Use" = rda_use,
         "Number of publications" = n)
rda_use
```

## How is PCA used?

PCA use level 1

```{r}
pca_use <-
  df_supp %>% 
  filter(!is.na(pca_use)) %>% 
  mutate(pca_use = fct_infreq(pca_use) %>% fct_relevel("PCs used as variables", after = Inf)) %>% 
  group_by(pca_use) %>% 
  count() %>% 
  rename("Use" = pca_use,
         "Number of publications" = n)
pca_use
```

If PCs are used as variables, how are they used?

```{r}
pc_variable_use <-
  df_supp %>% 
  filter(!is.na(pc_variable_use)) %>% 
  mutate(pc_variable_use = fct_infreq(pc_variable_use) %>% 
           fct_relevel("Interest in relationships with raw variables",
                       after = Inf)) %>% 
  group_by(pc_variable_use) %>% 
  count() %>% 
  rename("Use" = pc_variable_use,
         "Number of publications" = n)

pc_variable_use
#check that all pc uses accounted for
sum(pc_variable_use[2]) == pca_use[4, 2]
```

When PCs are used as variables in other models, do the studies have multivariate responses or multivariate predictors/explanatory variables?

```{r}
df_supp %>% 
  filter(pca_use == "PCs used as variables") %>% 
  count(mv_response_or_predictor)
```




If the authors are interested in doing a supervised type test, is there a more appropriate supervised drop-in replacement for what they did?

```{r}
supervised_alt <-
  df_supp %>% 
  filter(!is.na(supervised_alt)) %>% 
  mutate(supervised_alt = fct_relevel(supervised_alt, "yes", "no", "maybe")) %>% 
  group_by(supervised_alt) %>% 
  count()  %>% 
  rename("Use" = supervised_alt,
         "Number of publications" = n)
supervised_alt
```


Combine, export, do some editing in Excel or Numbers for publication 

```{r}
bind_rows(pls_use, rda_use, pca_use, pc_variable_use, supervised_alt) %>%
  write_csv(here("out", "review-table-rough.csv"))
```

## How often are other multivariate methods used?

```{r}
df_supp %>% 
  filter(!is.na(other_supervised)) %>% count()
```

## How manuy uses of each "other" supervised method, etc.

```{r}
df_supp %>% 
  filter(!is.na(other_supervised)) %>% 
  separate_rows(other_supervised, sep = ", ") %>% 
  count(other_supervised) %>% 
  arrange(desc(n))
```

## How is PERMANOVA and PCA used together?

```{r}
df_supp %>% 
  filter(!is.na(other_supervised)) %>% 
  separate_rows(other_supervised, sep = ", ") %>% 
  filter(other_supervised == "PERMANOVA") 
```



