---
title: "Additional Scenarios"
output: html_notebook
---

Here I'll simulate some additional scenarios for supplemental.

1. Needle in a haystack

Essentially a more extreme version of "hidden predictors". Many co-varying predictors, forming multiple axes of variation, that don't discriminate between groups, and only a few, uncorrelated, discriminating variables.

2. Red Herring
    - no discriminating variables
    - 5 variables with cov = 0
    - two groups of 10 variables with cov = 0.5
    
The idea here is that PCA-LR should find 2 principal components (on average), but neither will explain group membership.  PLS-DA should fail to find any significant components.

# Packages

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(holodeck)
library(ropls)
library(furrr)
source(here("R", "pcr.R"))
source(here("R", "ropls_helpers.R"))
source(here("R", "RMSEP.R"))
```

```{r message=FALSE, warning=FALSE}
plan("multisession")
```

```{r}
# these "safe" versions will never error, but return NULL if there is an error instead.
safe_opls <- safely(partial(opls, info.txtC = "none", fig.pdfC = "none"), otherwise = NULL)
safe_pcr <- safely(pca_lr, otherwise = NULL)
```

# Needle in a haystack

Let's try to get like 3 or 4 main axes of variation and then a sprinkling of actually discriminating variables.

```{r}
# runif(1, 1, 1000)
set.seed(152)
needle <- future_map(1:100, ~{
  sim_cat(n_obs = 20, n_groups = 2) %>% 
  group_by(group) %>% 
  sim_discr(n_vars = 5, var = 1, cov = 0, group_means = c(-1, 1), name = "D") %>% 
  sim_covar(n_vars = 10, var = 1, cov = 0.5, name = "C1") %>% 
  sim_covar(n_vars = 10, var = 1, cov = 0.5, name = "C2") %>% 
  sim_covar(n_vars = 10, var = 1, cov = 0.5, name = "C3") %>% 
  sim_covar(n_vars = 10, var = 1, cov = 0.5, name = "C4") %>% 
  sim_covar(n_vars = 5, var = 1, cov = 0, name = "N") %>% 
  ungroup()
}) %>% set_names(paste0("needle_in_haystack_", 1:100))
# needle[[1]]
```


## PCA-LR

```{r}
needle.pca <- 
  future_map(needle, ~safe_pcr(.x, -group, group)) %>%
  future_map("result") %>%
  discard(is.null)

get_loadings(needle.pca[[1]]$pca) %>%
  arrange(desc(abs(p1)))
```

## PLS-DA

```{r}
needle.pls <-
  future_map(needle, ~safe_opls(select(.x, -group), .x$group, permI = 200)) %>%
  future_map("result") %>%
  discard(is.null)
```

## Summary

```{r}
needle.pls.summary <- pls.summary(needle.pls)
needle.pcr.summary <- pcr.summary(needle.pca)
```

```{r}
needle.pcr.summary
needle.pls.summary
```

# Red Herring


```{r}
herring <- 
  future_map(1:500,
             ~sim_cat(n_obs = 20, n_groups = 2) %>% 
               sim_covar(n_vars = 5, var = 1, cov = 0, name = "N") %>% 
               sim_covar(n_vars = 10, var = 1, cov = 0.5, name = "C1") %>% 
               sim_covar(n_vars = 10, var = 1, cov = 0.5, name = "C2")
  ) %>% 
  set_names(paste0("herring_df_", 1:500))
```

## PCA-LR

```{r}
herring.pca <- 
  future_map(herring, ~safe_pcr(.x, -group, group)) %>% 
  future_map("result") %>%
  discard(is.null)

get_loadings(herring.pca[[1]]$pca) %>% arrange(desc(abs(p1)))
```

## PLS-DA

```{r}
herring.pls <-
  future_map(herring, ~safe_opls(select(.x, -group), .x$group, permI = 200)) %>%
  future_map("result") %>%
  discard(is.null)
```

## Summary

```{r}
herring.pls.summary <- pls.summary(herring.pls)
herring.pcr.summary <- pcr.summary(herring.pca)
```

```{r}
herring.pcr.summary
herring.pls.summary
```

The p-value % is misleading for PLS-DA since only 84 out of 500 even had 1 significant component.

```{r}
.48*84/500*100
```

It's really only 8% of the pQ2 < 0.05